---
title: "Titanic Data Set, BlackForest"
author: "Michelle Migdalovich"
output: html_document


---
## R Markdown
**Random Forest:**
A random forest was performed to ascertain how well certain variables were at predicting survival rates of the Titanic passengers. Random forest is convenient in that it is able to predict important variables on its own, however for this exercise the main variables were hand-picked.


**Data Set-Up & Cleaning**
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)


library(party)
library(rattle)
library(rpart.plot)
library(RColorBrewer)
library(caret)
library(ROCR)
library(caret)
library(lattice)
library(ggplot2)
library(rpart)
library(RColorBrewer)
library(rpart.plot)
library(rattle)
library(randomForest)
library(ROCR)
library(rpart.plot)
library(caret)
library(ggplot2)



# Read in Callie Test
source("C:/Users/mmigdalovich/Documents/Titanic/CallieTest.R")

```

```{r}
{include=TRUE}
knitr::opts_chunk$set(echo = FALSE)

# Read in Callie Test
source("C:/Users/mmigdalovich/Documents/Titanic/CallieTest.R")

#Read in data
titanicdata <- read.csv("C:/Users/mmigdalovich/Documents/train.csv")
summary(titanicdata)

titanicdata$Age[is.na(titanicdata$Age)] <- mean(titanicdata$Age,na.rm=T) 

#split the data using the function
y <- CallieTest(titanicdata,.50, TRUE) 
train<- data.frame(y[1])
test<-data.frame(y[2])

# Create levels
titanicdata$Pclass <- factor(titanicdata$Pclass, levels=c("3","2","1"), ordered=TRUE)
titanicdata$Age[is.na(titanicdata$Age)] <- mean(titanicdata$Age,na.rm=T) # filling in missing ages
```

This document displays a random forest of a dataset containing data describing the survivors of the Titanic tragedy. The variables used for analyses are Passanger Class (PClass), Age, and Sex.

Passenger Class (PClass): Class of passenger, with one being the 1st class being the highest and 3rd being the lowest

Age: Age of the Passenger. Not all ages were provided, and therefore missing data was filled using a mean data function

Sex: Sex of the passenger



```{r }
#My Forest
BlackForest <- randomForest(as.factor(Survived) ~ Pclass + Sex + Age, data=train, importance=TRUE, ntree=2000)


# view results 
print(BlackForest) 

# importance of each predictor
imp <- importance(BlackForest) 

#Subset of the Data
newdata <-subset(test, select=c(3, 5, 6)) 
My_PredictionRF<- data.frame(predict(BlackForest, newdata, type="class"))



```
##Evaluating Performance
A confusion matrix was generated to ascertain misclassification rate, ROC curves were calculated, as well as Area Under the Curve (AUC)
```{r}
#misclassification rate
confusionmat <- confusionMatrix(My_PredictionRF[[1]], test$Survived)  #confusion matrix
confusionmat
#ROC Curve
My_PredictionRF2<- data.frame(predict(BlackForest, newdata, type="prob"))
pr <- prediction(My_PredictionRF2[2], test$Survived)
prf <- performance(pr, measure = "tpr", x.measure = "fpr")
plot(prf)

#AUC
auc <- performance(pr, measure = "auc")
auc <- auc@y.values[[1]]
auc
```

## Plotting the Data

The Random Forest Data was plotted, displaying that **"sex"** is a very immportant factor when analyzing "survivors"



```{r}
vars<-dimnames(imp)[[1]]
imp<-data.frame(vars=vars,imp=as.numeric(imp[,1]))
imp<-imp[order(imp$imp,decreasing=T),]
par(mfrow=c(1,2))


varImpPlot(BlackForest,main='Variable Importance Plot: Base Model')
plot(BlackForest,main='Error vs No. of trees plot: Base Model')
```
```


