---
title: "Movie Lens Data Cleaning and Merging Demo"
author: "CZ"
date: "December 5, 2016"
output: html_document
---

### Project Overview


#### Assignment 
Using the Movie Lens data, we were instructed to clean and merge the data sets in order to: 

* Construct, clean, and label a data frame including the following variables: 
    + Demographic Info: user id, gender, age 
    + Movie Info: movie id, rating, time stamp, title, genre, year of movie 
    + Location Info: zip code, state, longitude, latitude 
* Generate a summary table of average reviews by gender for each gender. Visualize the results. 
* Create a heat map of the entire country of the count of reviews. 
* Create a csv file of the new data frame. 

***
 
#### Data Intro  
This exercise was designed to utilize publicly available Movie Lens data sets provided generously by the University of Minnesota. In total, four data sets were used. They are summarized in the table below.

<center>

Data set | Description                     | Key Variables
---------| ------------------------------- | -------------
u.data | The full u data set, 100,000 ratings by 943 users on 1682 items. Each user has rated at least 20 movies. Users and items are numbered consecutively from 1. The data is randomly ordered. | user id, item id, rating, time stamp
u.item | Information on the items (movies). | movie id, movie title, release date, video release date, IMDb URL, 19 genres
u.user | Demographic information about the users | user id, age, gender, occupation, zip code
zips.csv | Information on location. | zip code, state, longitude, latitude

</center>

***
 
####Document Sections
The following document will proceed as follows:
<b>

 1. Set up
 2. Cleaning the data 
 3. Merging the data frames
 4. Summary Statistics
 5. Heat Map
 6. Export to CSV
 
</b>

***

###Set Up
Set up includes installing and reading in the necessary packages and libraries. It also includes reading in the four data sets listed above. Packages used include:

* plyr
* dplyr
* tidyr
* stringer
* ggmap
* ggplot2
<br>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#R-Training
#Cleaning and Merging Datasets 
#Movie Lens Datasets
#11/30/16

############# SET UP ############# 
#Install packages if don't have them
#install.packages("plyr")
#install.packages("ggmap")
#install.packages("stringr")
#install.packages("dplyr")
#install.packages("tidyr")
#install.packages("varhandle")
#install.packages("data.table")
#install.packages("ggplot2")

#Load packages
library(party)
library(rpart)
library(rattle)
library(rpart)  
library(rpart.plot)
library(RColorBrewer)
library(caret)
library(lattice)
library(e1071)
library(pROC)
library(ROCR)
library(ggplot2)
library(tree)
library(randomForest)
library(gridExtra)
library(rmarkdown)
library(plyr)
library(ggmap)
library(stringr)
library(dplyr)
library(tidyr)
library(varhandle)
library(data.table)

############# LOADING THE DATA ############# *****************************NEED TO TAKE OUT FILE PATHS BEFORE UPLOADING!!

#Dataset 1: U.data contains user id, movie id, rating, timestamp
u_data <- read.delim("C:/Users/INPUT OWN PATH/Trainings/R_Studio/Data/u.data",
                        header = FALSE)
  
#Dataset 2: U.item contains movie id, movie title, release data, genre
u_item <- read.delim("C:/Users/INPUT OWN PATH/Trainings/R_Studio/Data/u.item",
                       header = FALSE, sep = "|")
  
#Dataset 3: U.user contains user id, age, gender, zip code
u_user <- read.delim("C:/Users/INPUT OWN PATH/Trainings/R_Studio/Data/u.user",
                       header = FALSE, sep = "|")
  
#Dataset 3: U.user contains user id, age, gender, zip code
zips_csv <- read.csv(
    "C:/Users/INPUT OWN PATH/Trainings/R_Studio/Data/zips.csv",
    header = TRUE
)

```

Summaries of the four data sets are below. As you can see, only the last data set, zips_csv has variable labels. We will add them in the next section, Data Cleaning. 

```{r datasets_summary}
#Dataset 1: U.data contains user id, movie id, rating, timestamp
summary(u_data)
  
#Dataset 2: U.item contains movie id, movie title, release data, genre
summary(u_item)
  
#Dataset 3: U.user contains user id, age, gender, zip code
summary(u_user)
  
#Dataset 3: U.user contains user id, age, gender, zip code
summary(zips_csv)
```

###Cleaning the Data
In order to prepare the data sets for merging, a series of data cleaning tasks needed to be carried out. 

1. Variable labels were manually added.
    a. Note: I assumed that the variable "item_id" from u.data was the same as "movie_id" in u.item. 
2. The date variable from u_item was split into day, month, year so we could access the movie release year.
3. Latitude and Longitude in zips_csv were converted to numeric variables from factor. 
    a. Note: In additional to using as.numeric, I also needed to remove the quotes around the numbers.
    b. Note: An area of concern is that there are many missing lat and long values. This will interfere with the accuracy of the heat map later.


```{r datacleaning, include= FALSE}
############# CLEANING THE DATA ############# 

#DATASET 1: U_Data
  #Adding in labels
  names(u_data) <- c("id","movie_id","rating", "timestamp")


#DATASET 2: U_Item
  #Adding in labels
  names(u_item) <- c("movie_id", "movie_title", "release_date", "video_release_date", "url", "unknown", "action",
                     "adventure", "animation", "children", "comedy", "crime", "documentary", "drama", "fantasy", "film_noir",
                     "horror", "musical", "mystery", "romance", "sci_fi", "thriller", "war", "western")
  str(u_item)

  #Splitting out the date to the year (last 4 digits)
    #Change date into a character
    u_item$date_chr <- as.character.Date(u_item$release_date)
   
    separate(u_item, date_chr, c("day", "month", "year"), "-", fill = "right")
    #Note: One missing date at row 267, filled in from the right. It remains blank.
  

#DATASET 3: U_User
  #Adding in labels
  names(u_user) <- c("id","age","gender", "ocupation", "zipcode")
  
#DATASET 4: zips_csv
  #Adding in labels
  names(zips_csv) <- c("zipcode", "state_abbrev", "lat", "long", "city", "state ") 
  
  #Change lat and long to numeric from factor
    #Remove quotation marks
    del <- colwise(function(zips_csv) str_replace_all(zips_csv, '"', ""))
    zips_csv2 <- del(zips_csv)
    
    #Chaning to numeric
    zips_csv2$lat <- as.numeric(zips_csv2$lat)
    zips_csv2$long <- as.numeric(zips_csv2$long)
```

###Merging Data Sets
After cleaning the data, I simply used the merge function to create one aggregate data frame, merge_all. 

```{r merging, echo=FALSE}

############# MERGING DATASETS ############# 
    
  #Merging u_data and u_item
  merge_temp1 <- merge(u_data,u_item,by="movie_id")
  
  #Merging in u_user
  merge_temp2 <- merge(merge_temp1,u_user,by="id")
  
  #Merging in zips_csv
  merge_all <- merge(merge_temp2, zips_csv2,by="zipcode", all.x=TRUE) 
    #Note: Need to left join so that unmatched ones are also added
  #merge_all
  
  #summary(merge_all)

```

###Summary Statistics
We were tasked with creating a summary table to display the average review by genre for each gender. This section turned out to be a lot trickier than I initially realized. 
<br> <br> 
<b> Complication: </b> A simple aggregate function does not work because the movie genre categories are not mutually exclusive. Because movies are labeled under multiple genres, the aggregate function splits up the genres into too many groups. Whereas I want an average of all movies under a certain genre say, drama, the aggregate function was returning different values for movies that were just drama, drama and action, and drama, action, and film noir, for example.
<br><br>
<b> Solution: </b> In order to address this issue, I created separate columns for each genre of the ratings assigned to that movie. If there was a 1 in the genre row, the associated rating for that movie would be carried over to the new genre rating column. For instance, if we were only looking at a section of the data frame (rating, action, and drama), we would create the two new columns on the right hand side. 

Rating | Action | Drama | Action Rating | Drama Rating
-------|--------|-------|---------------|-------------
2      |1       |0      |2              |0
3      |0       |1      |0              |3
3      |1       |1      |3              |3
4      |0       |0      |0              |0

<br>

After using a for loop to create these new rating variables, I was then able to use the aggregate function to create the table below. Following that is a bar graph of the same table.


```{r summary, echo=FALSE, warning=FALSE}
############# SUMMARY STATISTICS ############# 
  
#Summary table for display of average review by genre for each gender & Visualize
  
  #First attempt
  # Table3 <- with(merge_all, aggregate(merge_all$rating, list(gender, unknown, action),FUN=mean))
  # Table3
  
  #Note: Issue with summary table: movie label categories are not exclusive. 
    #Tables are counting 0,0, 0,1, 1,0, 1,1 as different groups if only looking at two genres
    #In actuality, want categories with any 1 to be summed up, regardless of multiple categories 
  #Solution: Create separate columns with drama rating, action rating, ect. and then find the means of those columns
    
  #Creating loop to create all the genre rating columns
    for(i in names(merge_all[,c(10:28)])){
      merge_all[[paste(i, 'r', sep="_")]] <- as.numeric(ifelse(merge_all[[i]]==1, merge_all$rating, "na"))
      #Note: Needed to put in na instead of 0's because didn't want 0 to be included in the mean. Ratings are only for 1-5.
      #Then needed to make sure casted it as a numeric so could calculate mean
    }
    
  #Summary check of newly created columns
  #summary(merge_all[,c(38:56)])
    
    
  #Aggregate table
  #Resource: http://stackoverflow.com/questions/21982987/mean-per-group-in-a-data-frame
  mean_table <- aggregate(merge_all[, 38:56], list(merge_all$gender), mean, na.rm = TRUE) 
  mean_table
 
  
  #Visualizing summary table in bar plot
    #Resources: Melt - https://www.r-bloggers.com/using-r-barplot-with-ggplot2/
      #Renaming labels: http://www.cookbook-r.com/Manipulating_data/Renaming_levels_of_a_factor/

    # Testing what kind of data frame can be plotted
    # df <- data.frame(genre = c("unknown_r","unknown_r", "action_r", "action_r", "animation_r", "animation_r"), 
    #                  gender = c("male", "female","male", "female","male", "female"), 
    #                  avg_rating =   c(3.1,3.5,3.5,3.5,3.6,3.8))
    # qplot(x=genre, y=avg_rating, fill=gender, data=df, geom="bar") +
    #   geom_bar(stat="identity")
    
    #Reshape the data so that genres become a column
    melted <- melt(mean_table, id.vars=c("Group.1"))
    
    #Clean up labels
    names(melted) <- c("gender","genre","avg_rating")
    
    #Clean up factor labels of genre so it doesn't have '_r' at the end when graphing
    levels(melted$genre) <- gsub("_r", "", levels(melted$genre))
    #levels(melted$genre)
   
    
 #Bar Plot
    ggplot(melted, aes(factor(genre), avg_rating, fill = gender, xlabel="Movie Genres")) + 
      geom_bar(stat="identity", position = "dodge") + 
      scale_fill_brewer(palette = "Set3") +
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
      xlab("Movie Genres") +
      ylab("Average Rating")+
      ggtitle("Average Movie Rating by Genre and Gender")

```



### Heat Map

The last visualization was a heat map of the number of ratings per regions. See below for the results.

```{r heatMap, echo=FALSE, warning = FALSE, error = FALSE}
############# HEAT MAP ############# 
#Heat map of entire country, by long and lat of count of reviews (more reviews = hotter)
#     #Resources:
#       #https://blog.dominodatalab.com/geographic-visualization-with-rs-ggmaps
    
    us_map <- get_map(location='united states', zoom=4, maptype = "roadmap",
                      source='google',color='color') 
    ggmap(us_map, extent = "device") + 
      geom_density2d(data = merge_all, aes(x = long, y = lat), size = 0.3) + 
      stat_density2d(data = merge_all, aes(x = long, y = lat, fill = ..level.., alpha = .95), 
                     size = 0.01,bins = 16, geom = "polygon") + scale_fill_gradient(low = "#e8f442", high = "#f44242") + 
      scale_alpha(range = c(0, 0.3), guide = FALSE)


```

###Export to CSV
This was the final part of the assignment. It was completed with a simple line of code:

<center> write.csv(merge_all, file = "Movie Lens_Merged_DF.csv") </center>

```{r export, echo=FALSE, warning = FALSE}
    
############# EXPORT TO CSV ############# 
    
#Set working directory
  setwd("C:/Users/INPUT OWN PATH/Trainings/R_Studio/Data") 

write.csv(merge_all, file = "MovieLens_Merged_DF.csv")
```

***

<center> 
###That's it! Thanks for reading. 
</center>
